Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2020-11-19 paper/ifo-introduction.tex
full grammar check before submission

% CW 
To date the network of gravitational wave observatories has observed short-duration transient signals from the mergers of binary black holes~\cite{GW150914,GW151226,GW170104,GW170814,GW190521,GWTC-1:2018,GWTC-2:2020} and binary neutron stars~\cite{GW170817,GW170817multi,GW190425,GWTC-1:2018,GWTC-2:2020}, as well as a black hole merger with an object that could either have been a neutron star or another black hole~\cite{GW190814}. 
However, the network is also searching for continuous gravitational waves: persistent, periodic, near-monochromatic signals, which are yet to be detected. 
Continuous waves may be emitted at a constant frequency or may wander slowly in frequency over time. 
Rotating neutron stars are prime candidates for continuous wave emission, especially those in low mass X-ray binaries (LMXB), where the neutron star is in orbit with a low mass stellar companion.
%where a compact object such as a black hole or a neutron star is in orbit with a low mass stellar companion.
% we're talking about neutron stars in LMXB's, so we don't need to mention black holes in LMXB's.
The rotation frequency of the neutron star can wander over time due to variable accretion of matter (and hence angular momentum transfer) from the stellar companion~\cite{xraybinaries:1997}. 
Scorpius X-1 is a prime LMXB target for continuous wave searches. 
%Scorpius X-1 is an LMXB which is particularly bright in X-rays, making it a prime target for continuous wave searches. %from LMXBs. 
Numerous, as yet unsuccessful, searches have been performed for Scorpius X-1 and other LMXBs~\cite{ScoX1O2Viterbi:2019, ScoX1ViterbiO1:2017, RadiometerO1O2:2019, SearchRadiometerO1:2017, SearchCrossCorrO1:2017, ScoX1ViterbiO1:2017, SearchTwoSpecS6:2017, MiddletonEtAlO2LMXBs:2020}.

% changed present to use, as this is not the first time a Michelson has been presented as a tpy GW detector - hannah
In this paper, we use a table-top Michelson interferometer as a toy gravitational wave detector designed to detect sound instead of gravitational waves. 
We then extend its use to an ``optical microphone'' (using light to capture sound) and present a range of example analysis techniques for educators to use. 
% re-added "as an undergraduate experiment to be clear we are not talking about school teaching - hannah
As an undergraduate lab experiment, the demonstration can be used to teach topics ranging from continuous-wave detection and analysis to electronics, signal processing, and speech enhancement. 
%% I don't like the use of ``freedom to explore''. 
% removed Additionally - I don't think it is neccessary - hannah
It allows students in courses such as physics and electrical engineering to explore the response of an accessible, yet nontrivial, opto-mechanical system using a hierarchy of data analysis techniques of increasing complexity, including those used in the search for continuous waves in LIGO and Virgo data~\cite{ScoX1O2Viterbi:2019, SuvorovaEtAl:2017,SuvorovaEtAl:2016}. 
% This demonstration can be adapted to an effective outreach tool to explain gravitational wave science to a non-specialist audience.
% Rephrased as this kind of sounds like we are an exhibition too [hannah] - This demonstration also has the potential to be used as an outreach tool alongside other exhibitions
%This demonstration also has the potential to be used as an outreach tool alongside a range of other public engagement demonstrations and activities including; hands-on demonstrations and exhibitions~\cite{TTExhibit:2020,CavagliaExhibit:2009,LIGOScienceEducationCentre:online,GravityDiscoveryCentre:online}, online data analysis tutorials~\cite{GWOSC:online,LOSC:2015}, phone applications~\cite{LaserLabs:online,SciVR:online,chirp:online}, online games~\cite{BlackHoleHunter:online}, and musical performances~\cite{ArthurJeffesMusic:online,GravitySynthLeonTrimble:online} which have been developed by gravitational wave research groups around the world. 
This demonstration also has the potential to be used as an outreach tool alongside a range of other public engagement demonstrations and activities developed by gravitational wave research groups around the world~\cite{TTExhibit:2020,CavagliaExhibit:2009, LIGOScienceEducationCentre:online, GravityDiscoveryCentre:online, GWOSC:online, LOSC:2015, LaserLabs:online, SciVR:online,chirp:online, BlackHoleHunter:online, ArthurJeffesMusic:online, GravitySynthLeonTrimble:online}. 
These tools allow scientists to cater to the increased public and media interest in this field and explain gravitational wave science to non-specialist audiences.
%\han{look up Sarah Farmer's project}

>>>>

% CW 
To date, the network of gravitational wave observatories has observed short-duration transient signals from the mergers of binary black holes~\cite{GW150914,GW151226,GW170104,GW170814,GW190521,GWTC-1:2018,GWTC-2:2020} and binary neutron stars~\cite{GW170817,GW170817multi,GW190425,GWTC-1:2018,GWTC-2:2020}, as well as a black hole merger with an object that could either have been a neutron star or another black hole~\cite{GW190814}. 
However, the network is also searching for continuous gravitational waves: persistent, periodic, near-monochromatic signals, which are yet to be detected. 
Continuous waves may be emitted at a constant frequency or may wander slowly in frequency over time. 
Rotating neutron stars are prime candidates for continuous wave emission, especially those in low mass X-ray binaries (LMXB), where the neutron star is in orbit with a low mass stellar companion.
%where a compact object such as a black hole or a neutron star is in orbit with a low mass stellar companion.
% we're talking about neutron stars in LMXB's, so we don't need to mention black holes in LMXB's.
The rotation frequency of the neutron star can wander over time due to variable accretion of matter (and hence angular momentum transfer) from the stellar companion~\cite{xraybinaries:1997}. 
Scorpius X-1 is a prime LMXB target for continuous wave searches. 
%Scorpius X-1 is an LMXB which is particularly bright in X-rays, making it a prime target for continuous wave searches. %from LMXBs. 
Numerous, as yet unsuccessful, searches have been performed for Scorpius X-1 and other LMXBs~\cite{ScoX1O2Viterbi:2019, ScoX1ViterbiO1:2017, RadiometerO1O2:2019, SearchRadiometerO1:2017, SearchCrossCorrO1:2017, ScoX1ViterbiO1:2017, SearchTwoSpecS6:2017, MiddletonEtAlO2LMXBs:2020}.

% changed present to use, as this is not the first time a Michelson has been presented as a tpy GW detector - hannah
In this paper, we use a table-top Michelson interferometer as a toy gravitational wave detector designed to detect sound instead of gravitational waves. 
We then extend its use to an ``optical microphone'' (using light to capture sound) and present a range of example analysis techniques for educators to use. 
% re-added "as an undergraduate experiment to be clear we are not talking about school teaching - hannah
As an undergraduate lab experiment, the demonstration can be used to teach topics ranging from continuous-wave detection and analysis to electronics, signal processing, and speech enhancement. 
%% I don't like the use of ``freedom to explore''. 
% removed Additionally - I don't think it is neccessary - hannah
It allows students in courses such as physics and electrical engineering to explore the response of an accessible, yet nontrivial, optomechanical system using a hierarchy of data analysis techniques of increasing complexity, including those used in the search for continuous waves in LIGO and Virgo data~\cite{ScoX1O2Viterbi:2019, SuvorovaEtAl:2017,SuvorovaEtAl:2016}. 
% This demonstration can be adapted to an effective outreach tool to explain gravitational wave science to a non-specialist audience.
% Rephrased as this kind of sounds like we are an exhibition too [hannah] - This demonstration also has the potential to be used as an outreach tool alongside other exhibitions
%This demonstration also has the potential to be used as an outreach tool alongside a range of other public engagement demonstrations and activities including; hands-on demonstrations and exhibitions~\cite{TTExhibit:2020,CavagliaExhibit:2009,LIGOScienceEducationCentre:online,GravityDiscoveryCentre:online}, online data analysis tutorials~\cite{GWOSC:online,LOSC:2015}, phone applications~\cite{LaserLabs:online,SciVR:online,chirp:online}, online games~\cite{BlackHoleHunter:online}, and musical performances~\cite{ArthurJeffesMusic:online,GravitySynthLeonTrimble:online} which have been developed by gravitational wave research groups around the world. 
This demonstration also has the potential to be used as an outreach tool alongside a range of other public engagement demonstrations and activities developed by gravitational wave research groups around the world~\cite{TTExhibit:2020,CavagliaExhibit:2009, LIGOScienceEducationCentre:online, GravityDiscoveryCentre:online, GWOSC:online, LOSC:2015, LaserLabs:online, SciVR:online,chirp:online, BlackHoleHunter:online, ArthurJeffesMusic:online, GravitySynthLeonTrimble:online}. 
These tools allow scientists to cater to the increased public and media interest in this field and explain gravitational wave science to non-specialist audiences.
%\han{look up Sarah Farmer's project}

### 2020-11-19 paper/ifo-tableTopGWs.tex
full grammar check before submission

% rephrased to mention the laser as the first component and shifted the arm lengths of LIGO and Virgo to later in the para when we discuss them. 
Gravitational wave detectors such as LIGO and Virgo are large, complex experiments. 
However their design is fundamentally based on the Michelson interferometer, an optical configuration commonly used in undergraduate laboratories. 
In a Michelson interferometer, laser light is split by a beamsplitter into two perpendicular arms (see Fig.~\ref{fig:ifo_schematic_webcam}). 
Mirrors at the end of each arm reflect the two beams back to the beamsplitter where they recombine to produce an interference pattern, 
The resulting interference pattern is dependent on the relative distance travelled by the beams. 
Current generation gravitational wave observatories have kilometre-scale arms (with arm lengths $4\,{\rm km}$ and $3\,{\rm km}$ at LIGO and Virgo respectively). 
They are able to measure minuscule changes in distance due to gravitational waves, e.g. the first detection of a binary black hole merger produced a strain of $10^{-21}$, corresponding to a change in distance which is a fraction of the width of a proton. 
%changed the relative length of the two arms by $10^{-21}$, a distance smaller than the width of a proton~\cite{GW150914}.

>>>>

% rephrased to mention the laser as the first component and shifted the arm lengths of LIGO and Virgo to later in the para when we discuss them. 
Gravitational wave detectors such as LIGO and Virgo are large, complex experiments. 
However, their design is fundamentally based on the Michelson interferometer, an optical configuration commonly used in undergraduate laboratories. 
In a Michelson interferometer, laser light is split by a beamsplitter into two perpendicular arms (see Fig.~\ref{fig:ifo_schematic_webcam}). 
Mirrors at the end of each arm reflect the two beams back to the beamsplitter where they recombine to produce an interference pattern, 
The resulting interference pattern is dependent on the relative distance travelled by the beams. 
Current generation gravitational wave observatories have kilometre-scale arms (with arm lengths $4\,{\rm km}$ and $3\,{\rm km}$ at LIGO and Virgo respectively). 
They can measure minuscule changes in distance due to gravitational waves, e.g. the first detection of a binary black hole merger produced a strain of $10^{-21}$, corresponding to a change in distance which is a fraction of the width of a proton. 
%changed the relative length of the two arms by $10^{-21}$, a distance smaller than the width of a proton~\cite{GW150914}.

### 2020-11-19 paper/ifo-wanderingTone.tex
full grammar check before submission

A ``detection statistic'' is used to quantify the likelihood of a signal being present in the data at each frequency and timeseries segment. 
In gravitational wave data analysis, the detection statistic gives the likelihood of a signal given the antenna beam pattern of the detector, which varies as the Earth rotates and orbits the Sun~\cite{JKS:1998}. 
%Most continuous wave searches use detection statistics to quantify the likelihood of a signal being present in the data at each frequency for each timeseries segment. 
%The detection statistic used depends on the antenna beam pattern of the detector, which varies as the Earth rotates and orbits the Sun, and the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}.
The choice of detection statistic used depends on the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}.
In this work, we take the discrete Fourier transform of each timeseries segment and use its amplitude at each frequency as a detection statistic.
This maximises the likelihood of detecting a sinusoidal signal in Gaussian noise, as described in Appendix~\ref{app:sinusoid_likelihood}. 
The detection statistic is normalised for convenience and we use it to form a grid in time and frequency (a spectrogram) which displays the frequency content of the signal over time. 
% moved one sentence up 
%We also normalise the detection statistics for convenience.

@@

\begin{figure}
\includegraphics[width=0.49\textwidth]{figures/viterbiDiagram.pdf}
\caption{\label{fig:viterbi}
A schematic diagram of the Viterbi algorithm. 
Each circle represents an element in the time-frequency grid starting from the instant $t_0$ on the left and ending at $t_{N_t}$ on the right. 
Vertically, the grid starts from frequency $f_0$ at the bottom and ending at $f_{N_f}$ at the top. 
The colour of each circle indicates the likelihood that a signal is present in the box, where a lighter colour corresponds to higher likelihood. 
At each time, the paths can move up one $f$ bin, move down one $f$ bin, or stay at the same $f$ bin (shown by lines in the diagram). 
Each circle has three paths leading to it, indicated by the lines. 
The best path to each circle at each $t_i$ is highlighted in black. 
Some routes through the grid are found to be dead ends, with respect to optimally reaching the other side, such as the path ending at $t_1$ marked (a). 
At $t=t_{N_t}$ the algorithm chooses the terminating frequency circle which has the highest value given by Eqn.~\ref{eqn:recursiveViterbi}, marked (b). 
The Viterbi path is the path leading to this circle, highlighted in orange from (b) to the start at (c). 
}
\end{figure}

@@

The results are shown in Fig.~\ref{fig:viterbi_overlay}, where the heatmap shows the spectrogram of the observed signal. 
For the purposes of this demonstration we use a signal that can easily be identified in the spectrogram, however, we expect real continuous wave signals to have far lower signal-to-noise ratios (SNR's). 
The overlaid pink dots in Fig.~\ref{fig:viterbi_overlay} show the injected signal and the white crosses show the recovered Viterbi path.
% I think this is already defined and seems like repetition:(the most likely path through the data according to the Viterbi algorithm). 
The recovered Viterbi path is within one frequency bin ($\approx 0.3\,{\rm Hz}$) of the injected signal for $94\%$ of the time. We also compute the root-mean-square (RMS) fractional error $E_{\rm rms}$ along the path, which is defined as 
\begin{equation}
E_{\rm rms} = \left[ \frac{1}{N_t+1} \sum_{i=0}^{N_t} \frac{(I_i - R_i)^2}{{I_i}^2} \right] ^{1/2}
\end{equation}
where $I_i$ and $R_i$ are the injected and recovered (Viterbi) frequency paths. Computing this for the result shown in Fig.~\ref{fig:viterbi_overlay}, we find that $E_{\rm rms} = 0.082$, which indicates fair but not total recovery of the injected signal.

This may be explained by two anomalies in the recovered path. Initially, the injected signal wanders by more than one frequency bin per time bin (i.e.\ faster than the algorithm is allowed to), thus leading to a discrepancy between the injected and recovered paths for $t\lesssim 10\,{\rm s}$. One may be tempted to increase the allowed frequency wander in the algorithm, however this leads to an overall decrease in the above statistics, as the algorithm is prone to jump briefly to nearby spots of noise. There is also an anomaly at $150\,{\rm s}$, which is likely due to a local disturbance, e.g. someone walking past the interferometer. 
As shown in Fig.~\ref{fig:viterbi_overlay}, the Viterbi algorithm is able to recover and continues to track the signal after the disturbance. 
%The Viterbi algorithm is somewhat robust against such disturbances, qualitatively recovering the injected path on the other side, as shown in Fig.~\ref{fig:viterbi_overlay}.

>>>>

A ``detection statistic'' is used to quantify the likelihood of a signal being present in the data at each frequency and time series segment. 
In gravitational wave data analysis, the detection statistic gives the likelihood of a signal given the antenna beam pattern of the detector, which varies as the Earth rotates and orbits the Sun~\cite{JKS:1998}. 
%Most continuous wave searches use detection statistics to quantify the likelihood of a signal being present in the data at each frequency for each timeseries segment. 
%The detection statistic used depends on the antenna beam pattern of the detector, which varies as the Earth rotates and orbits the Sun, and the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}.
The choice of detection statistic used depends on the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}.
In this work, we take the discrete Fourier transform of each time series segment and use its amplitude at each frequency as a detection statistic.
This maximises the likelihood of detecting a sinusoidal signal in Gaussian noise, as described in Appendix~\ref{app:sinusoid_likelihood}. 
The detection statistic is normalised for convenience and we use it to form a grid in time and frequency (a spectrogram) which displays the frequency content of the signal over time. 
% moved one sentence up 
%We also normalise the detection statistics for convenience.

@@

\begin{figure}
\includegraphics[width=0.49\textwidth]{figures/viterbiDiagram.pdf}
\caption{\label{fig:viterbi}
A schematic diagram of the Viterbi algorithm. 
Each circle represents an element in the time-frequency grid starting from the instant $t_0$ on the left and ending at $t_{N_t}$ on the right. 
Vertically, the grid starts from frequency $f_0$ at the bottom and ending at $f_{N_f}$ at the top. 
The colour of each circle indicates the likelihood that a signal is present in the box, where a lighter colour corresponds to a higher likelihood. 
At each time, the paths can move up one $f$ bin, move down one $f$ bin, or stay at the same $f$ bin (shown by lines in the diagram). 
Each circle has three paths leading to it, indicated by the lines. 
The best path to each circle at each $t_i$ is highlighted in black. 
Some routes through the grid are found to be dead ends, with respect to optimally reaching the other side, such as the path ending at $t_1$ marked (a). 
At $t=t_{N_t}$ the algorithm chooses the terminating frequency circle which has the highest value given by Eqn.~\ref{eqn:recursiveViterbi}, marked (b). 
The Viterbi path is the path leading to this circle, highlighted in orange from (b) to the start at (c). 
}
\end{figure}

@@

The results are shown in Fig.~\ref{fig:viterbi_overlay}, where the heatmap shows the spectrogram of the observed signal. 
For this demonstration we use a signal that can easily be identified in the spectrogram, however, we expect real continuous wave signals to have far lower signal-to-noise ratios (SNR's). 
The overlaid pink dots in Fig.~\ref{fig:viterbi_overlay} show the injected signal and the white crosses show the recovered Viterbi path.
% I think this is already defined and seems like repetition:(the most likely path through the data according to the Viterbi algorithm). 
The recovered Viterbi path is within one frequency bin ($\approx 0.3\,{\rm Hz}$) of the injected signal for $94\%$ of the time. We also compute the root-mean-square (RMS) fractional error $E_{\rm rms}$ along the path, which is defined as 
\begin{equation}
E_{\rm rms} = \left[ \frac{1}{N_t+1} \sum_{i=0}^{N_t} \frac{(I_i - R_i)^2}{{I_i}^2} \right] ^{1/2}
\end{equation}
where $I_i$ and $R_i$ are the injected and recovered (Viterbi) frequency paths. Computing this for the result shown in Fig.~\ref{fig:viterbi_overlay}, we find that $E_{\rm rms} = 0.082$, which indicates fair but not total recovery of the injected signal.

This may be explained by two anomalies in the recovered path. Initially, the injected signal wanders by more than one frequency bin per time bin (i.e.\ faster than the algorithm is allowed to), thus leading to a discrepancy between the injected and recovered paths for $t\lesssim 10\,{\rm s}$. One may be tempted to increase the allowed frequency wander in the algorithm, however, this leads to an overall decrease in the above statistics, as the algorithm is prone to jump briefly to nearby spots of noise. There is also an anomaly at $150\,{\rm s}$, which is likely due to a local disturbance, e.g. someone walking past the interferometer. 
As shown in Fig.~\ref{fig:viterbi_overlay}, the Viterbi algorithm can recover and continues to track the signal after the disturbance. 
%The Viterbi algorithm is somewhat robust against such disturbances, qualitatively recovering the injected path on the other side, as shown in Fig.~\ref{fig:viterbi_overlay}.

### 2020-11-19 paper/paper_main.tex
full grammar check before submission

\begin{abstract}
Gravitational wave observatories around the world are searching for continuous waves: persistent signals from sources such as spinning neutron stars. 
% \jam{For the hyphen in "gravitational wave" versus "gravitational wave". I prefer the former, but either way, we should be consistent.}
These searches use sophisticated statistical techniques to look for weak signals in noisy data. 
In this paper, we demonstrate these techniques using a table-top model gravitational wave detector: a Michelson interferometer where sound is used as an analogue for gravitational waves. 
Using signal processing techniques from continuous-wave searches we demonstrate the recovery of tones with constant and wandering frequencies. 
We explore use of the interferometer as a teaching tool for educators in physics and electrical undergraduate laboratories with a range of example activities.
% \jam{We mention undergraduate labs twice, roll this sentence into the end? Also, electrical sounds weird, why the change from electrical engineering?}
The interferometer is used as an ``optical microphone'' to capture music and speech. 
% \jam{I changed all the \emph to ``'', feel free to change them back, but I think it is more explicit this way}
A selection of filtering techniques are applied to recover the original signals from noisy data, including simple bandpass filtering, notch filters, a Weiner filter, and the statistical logMMSE method.
Recordings of simple chords and drums are easily recovered, but complex music and speech are more challenging. 
This demonstration has scope for use by undergraduate laboratories and can be adapted for use as an engagement tool for communicating gravitational wave and signal-processing topics to non-specialist audiences. 
% rearranged in response to Anna's pnp review
%Finally we use the interferometer as an `optical microphone' to capture music and speech and apply filtering techniques to recover the original signals from the noisy data. Such techniques include simple bandpass filtering, notch filters, a Weiner filter, and the statistical logMMSE method.  
%Simple chords and drums are easily recovered, but complex music and speech are more challenging. 
%This demonstration has scope for use in physics and electrical engineering undergraduate laboratories. 
%It can also be adapted for use as an engagement tool for communicating gravitational wave and signal-processing topics to non-specialist audiences. 
\end{abstract}

>>>>

\begin{abstract}
Gravitational wave observatories around the world are searching for continuous waves: persistent signals from sources such as spinning neutron stars. 
% \jam{For the hyphen in "gravitational wave" versus "gravitational-wave". I prefer the former, but either way, we should be consistent.}
These searches use sophisticated statistical techniques to look for weak signals in noisy data. 
In this paper, we demonstrate these techniques using a table-top model gravitational wave detector: a Michelson interferometer where sound is used as an analogue for gravitational waves. 
Using signal processing techniques from continuous-wave searches we demonstrate the recovery of tones with constant and wandering frequencies. 
We explore the use of the interferometer as a teaching tool for educators in physics and electrical undergraduate laboratories with a range of example activities.
% \jam{We mention undergraduate labs twice, roll this sentence into the end? Also, electrical sounds weird, why the change from electrical engineering?}
The interferometer is used as an ``optical microphone'' to capture music and speech. 
% \jam{I changed all the \emph to ``'', feel free to change them back, but I think it is more explicit this way}
A selection of filtering techniques is applied to recover the original signals from noisy data, including simple bandpass filtering, notch filters, a Weiner filter, and the statistical logMMSE method.
Recordings of simple chords and drums are easily recovered, but complex music and speech are more challenging. 
This demonstration has scope for use by undergraduate laboratories and can be adapted for use as an engagement tool for communicating gravitational wave and signal-processing topics to non-specialist audiences. 
% rearranged in response to Anna's pnp review
%Finally we use the interferometer as an `optical microphone' to capture music and speech and apply filtering techniques to recover the original signals from the noisy data. Such techniques include simple bandpass filtering, notch filters, a Weiner filter, and the statistical logMMSE method.  
%Simple chords and drums are easily recovered, but complex music and speech are more challenging. 
%This demonstration has scope for use in physics and electrical engineering undergraduate laboratories. 
%It can also be adapted for use as an engagement tool for communicating gravitational wave and signal-processing topics to non-specialist audiences. 
\end{abstract}
